---
title: "introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r example}
library(iwillsurvive)
library(dplyr)
```

It's best to start with one-row-per-patient (ORPP) cohort object that contains 
columns corresponding to 

- `patientid`, a unique patient identifier
- `index_date`, a date corresponding to an index date.
- `censor_date`, date corresponding to when patients were censored
- `event_date`, date corresponding to the event of interest. NA values indicate 
that the event was not observed.

`iwillsurvive` provides one such example in `ez_cohort`, a dataframe of `r scales::comma(nrow(ez_cohort))`
simulated patients:

```{r}
cohort <- ez_cohort
```

# `derive_*()` functions

Use the `derive_*()` functions to calculate key derived columns:

`derive_followup_date()` returns the column `follow_up` date, which is event_date
 (if known), and the `censor_date` otherwise:

```{r}
cohort <- cohort %>%
  derive_followup_date(event_date = "dateofdeath",
                       censor_date = "lastvisitdate")

cohort
```

`derive_followup_time()` returns the columns followup_days, followup_months, and
followup_years which show time from index to follow up:

```{r}
cohort <- cohort %>%
  derive_followup_time(index_date = "lotstartdate") 

cohort %>%
  select(patientid, lotstartdate, followup_date, 
         followup_days, followup_months, followup_years)
```

`derive_event_status()` will add an `event_status` column

```{r}
cohort <- cohort %>%
  derive_event_status(event_date = "dateofdeath") 

cohort
```

