---
title: "introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(iwillsurvive)
```


Imagine you have the following dataframe

```{r}
ez_cohort
```


It's best to start with one-row-per-patient (ORPP) cohort object.

For this example, we'll `ez_cohort`, a dataframe of simulated patients contained in `iwillsurvive`:

```{r}
ez_cohort
```


Use `derive

```{r}
ez_cohort <- ez_cohort %>%
  derive_followup_date(event_date = "dateofdeath",
                        censor_date = "lastvisitdate") %>%
  
  derive_followup_time(index_date = "lotstartdate") %>%
  
  derive_event_status(event_date = "dateofdeath")
```




Use `plot_followup_time` to visualize the time at risk data

```{r, fig.width = 7, fig.height = 4, out.width = "85%"}
plot_followup_time(ez_cohort, 
                   followup_time = "followup_days", 
                   event_status = "event_status")
```

Use `fit_survival()` to fit the survival model. You can do this in two ways:

If you already have time at risk and the event status calculated (as we do in the `ez_cohort` object), you can simply point the function to those columns:

```{r}
cohort_fit <- fit_survival(ez_cohort, 
                           followup_time = "followup_days", 
                           event_status = "event_status", 
                           terms = "condition")
```

You should get the same result using both methods.

The result is a `survfit` object (from the `survival` package)

```{r}
class(cohort_fit)

cohort_fit
```

Use `plot_survival()` to plot the result.

```{r, fig.width = 7, fig.height = 5, out.width = "85%"}
plot_survival(cohort_fit, 
              cohort = ez_cohort)
```

# derive_*() functions

Placeholder for in depth descriptions of derive functions

# plot_*() functions

Placeholder for in depth descriptions of plot functions

# filter_*() functions

Placeholder for in depth descriptions of filter functions



